Detection of Arabic Entity Graphs using Morphology, Finite State Machines, and Graph Transformations

### 2011-11-24 tex/cicling.tex
updated code of biography segmenter (with 2 algorithms: the practical one is using lookups, the other is a true segmentation algorithm)

\usepackage{arabtex}
\usepackage{amssymb}
\usepackage{caption}
\usepackage{epsfig}
\usepackage{subfigure}
\usepackage{color}
\usepackage{rotate}
\usepackage{rotating}
%\usepackage{amsthm}
\usepackage{booktabs}

\usepackage{relsize}
\usepackage{fancyvrb}
\usepackage[colorlinks=false,draft]{hyperref}

\usepackage{utf8}
\setarab
\fullvocalize
\transtrue
\arabtrue

%\newcommand{\CharCodeIn}[1]{`\CodeIn{#1}'}
\newcommand{\CodeIn}[1]{{\small\texttt{#1}}}
\newcommand{\frl}[1]{\fbox{\RL{#1}}} 
\newcommand{\noArRL}[1]{\arabfalse\RL{#1}\arabtrue} 
\newcommand{\noTrRL}[1]{\transfalse\RL{#1}\transtrue} 
\newcommand{\noTrnoVocRL}[1]{\transfalse\novocalize\noTrRL{#1}\vocalize\transtrue}  
\newcommand{\noVocRL}[1]{\transtrue\novocalize\RL{#1}\vocalize}

%\title{Instructions for NAACL HLT 2010 Proceedings\Thanks{This...}}
\title{Knowledge-based Detection of Arabic Entity Graphs using Morphology and Graph Transformations}

%\author{ Jad Makhlouta \\\And
%Hamza Harkous \\
%  American University of Beirut \\
%  {\tt \{jem04, hhh20, fz11\}@aub.edu.lb } \\\And 
%  Fadi Zaraket 
%}

\date{}

\begin{document}

\maketitle

\begin{abstract}
A hadith in Islamic literature is a narration from the 
prophet Mohammad related by multiple narrators.
Establishing the authenticity of a hadith is an important task
based on which Islamic scholars issue rules that affect the life
of people around the world. 
The literature lacks an automated and exhaustive mechanism of 
authentication checks. 
This is currently manual work prone to error and almost
impossible to complete in the life time of a scholar due to the
large literature available. 
In this paper, we present a tool that uses 
an Arabic morphological analyzer 
to successfully automate the
analysis of three books of hadith selected 
arbitrarily and abstract 
each book into a graph of narrators related to each other. 
\end{abstract}

\section{Introduction}
 Arabic morphological rich
 entity detection in Arabic
 entity graphs
 locality of entities enables detection of entities and aggregating local clusters of entities into general graphs
 applications
 	bible genealogy 
 	hadith
 	bibliography
	travel itineraries
	others...
 learning and enriching gazetteers

\section{The Method}

Our method takes a set of documents and builds 
a graph where entities are nodes and edges are relations between the nodes. 
Our target graph has a predefined set of relations expressed as edge labels such as ``child'' and ``spouse''.
We use morphological analysis with finite-state machines to detect local entities and
relations between them.
We then use graph merging and comparison techniques to build the final general graphs.

In the sequel, we automatically learn entities and append them to our lexicon as new tags. 
We automatically learn lexicon items and infer their semantical meanings as relations in the 
graph merging phase. We express the inferred semantic meaning as a sequence of graph transformations
that removes the edge with the lexicon item and replaces it with edges with equivalent predefined labels.
We also interactively learn the semantics of other lexicon items that we detect as relations but could not infer their meaning 
in terms of the predefined set of relations.
The entities and relations are detected

Our method has two folds. At the first level, we detect entities and edges that
Given a lexicon that can distinguish entities and connections between them, and several documents that contain these entities,
we hypothesize that entities that occur next to each other are relevant entities even if their connections do not belong to the
connection lexicon. We also hypothesize that entities occur after connection are relevant entities.

-importance of cluster (Min number of entities in context)
-importance of tolerance 
-entities using lexicons
-clusters
-

\section{Application to Hadith}

A \RL{.hady_t}~\footnote{In this document, 
we use the default ArabTeX transliteration style ZDMG.}
(tradition) 
is a narration related to the prophet Mohammad
through a \RL{sanad} or a sequence of narrators. 
The collections of traditions are the second source of
jurisprudence after the \RL{qor'An} for all Islamic schools of thought. 
Figure~\ref{f:exhadith} shows an example \noArRL{.hady_t} in 
Arabic with its transliteration and translation. 
We show proper names in boxes connected
to form complex names of narrators. 
For example, 
\novocalize
\noTrRL{qtybT} is the first name 
of narrator $n_1$, and 
\noTrRL{s`yd} is the name of his father as 
the word \noTrRL{bn} (son of) indicates. 
The sequence of names from $n_1$ to $n_5$ 
constitutes the \noArRL{sanad} 
of the \noArRL{.hady_t}. 
The second part of the \noArRL{.hady_t} is the 
\RL{matn} (content) and constitutes the actual content 
of the tradition.

Due to religious and political reasons, 
writing the traditions was forbidden 
until the days of the eighth Umayad Calif,
\novocalize
\RL{`mr bn `bd al`zyz}(717-720 AC), seventy or so years after 
the death of the prophet~\cite{AlAskari}. 
Consequently, many inconsistencies were
introduced to the literature which necessitated 
a thorough authentication study of a tradition before its use in 
jurisprudence.
\vocalize
%An Islamic hadith scholar studies the authenticity of 
%the \noArRL{sanad} of a 
%set of related traditions before using these traditions 
%jurisprudence. 
While different Islamic schools of thought differ on 
how to interpret the content, they almost all agree
that if a \noArRL{sanad} lacks authenticity, 
then scholars can not use the tradition for jurisprudence.

The authenticity of a \noArRL{.hady_t} depends on 
the credibility of the narrators as reported in 
separate biography books. 
The study of \noArRL{.hady_t} authentication is 
currently manual and error prone due to the huge number
of existing traditions and tradition books. 
Hadithopedia~\cite{Hadithopaedia:08}
reports that the tradition and biography
books for one of the Islamic sects amounts to more than 
300 thousand lines of text. 
Al-Azami\cite{Al-Azami-91} cites more than eleven books
of digitized tradition books each of several volumes, and a dozen
other biography and secondary authentication books such
as a geographical dictionary of places in hadith.

@@

\transfalse
\begin{figure}[tb]
\center{
\resizebox{.9\columnwidth}{!}
{ \input{figs/exhadith.pdftex_t}}
\caption{Hadith abstraction example.}
\label{f:exhadith}
}
\end{figure}
% the FSMs for the three words
\transtrue
\vocalize

In this paper, we present the hadith {\em narrator graph
extractor using morphological analysis} (NaGEMA), 
a novel technique that automates
extracting the \noArRL{sanad} from the books of traditions 
into a diagram of chains of narrators. NaGEMA segments
a book into narrations, each narration into its \noArRL{matn} 
and its \noArRL{sanad}, 
and each sanad into its separate narrators. 
This work will lead to 
automate an exhaustive \noArRL{.hady_t} authentication 
effort in the future as we plan to 
automatically analyze the biography books.

%Automated analysis of Arabic data sets, including texts, 
%publications, records and digital media is essential
%with the huge digital Arabic content available nowadays. 
\subsection{ Arabic morphological analysis}
Arabic morphological analysis is key to our analysis. 
Current morphological analyzers~\cite{Sughaiyer:04}
use concatenative analysis when
considering the internal structure 
of an Arabic word and
composing it into several {\em morphemes}. 
A morpheme can be a {\em stem}, or an {\em affix}.
An affix can be a {\em prefix, suffix, } or an {\em infix}.
%The analysis of one word may lead to several possible
%morphological solutions.
\vocalize
The word \RL{'a.hmadH}
may have two valid morphological analyses. 
The letter \RL{'a} may be a prefix and the word means 
``I praise him'', or 
%The letter \RL{'a} may also be 
part of the stem \RL{'a.hmad} (a proper noun)
and the word means ``his Ahmad''.

The morphological analysis of the word \RL{.hd_tnA} 
``narrated to us'' that starts the hadith in Figure~\ref{f:exhadith}
returns the stem \RL{.hd_t} ``narrated''.
The stem  \RL{.hd_t} is also the stem of other 
words such as \RL{.hd_tny} ``narrated to me'' and \RL{.hd_thm} ``narrated to them''.
The stem also shares similar part of speech (POS)
and English meaning gloss annotations with words
such as \RL{qaal} ``said'' and \RL{'a_hbar} ``told''. 
The use of morphological analysis allows us to 
subsume listing these stems and all of their morphological
derivations with one rule that matches the gloss and 
POS tags. 
Similarly the words \RL{ibn} ``the son of'' and \RL{bn} 
return the same morphological analysis.

\novocalize
%The accuracy of the solutions suffer due to inherent difficulties
%of morphological analysis of the Arabic language. 
%For example, it is common practice to write Arabic text
%without short vowels. 
%This greatly increases the ambiguity of Arabic text. 
%Arabic letters can have up to 
%four different forms
%corresponding to their position in a word, i.e, beginning,
%middle, end of word and separate forms. 
%This allows the phrase \transfalse
%\RL{il_A\nospace almdrsT} \transtrue
%to be visually recognizable
%as two separate words \RL{il_A} (to) and \RL{almdrsT} (the school) 
%without the need of a space in between. 
%The reason is the first word \RL{il_A} ends with
%\RL{_A} a non-connecting letter. 
%These words,
%referred to as ``run-on'' words~\cite{Buckwalter:04},
%occur often, and greatly increase the
%difficulty of tokenization.

Current morphological analyzers such as 
Buckwalter~\cite{Buckwalter:02},
Beesley~\cite{Beesley:01},
SAMA~\cite{Kulick:10},
and ElixirFM~\cite{Otakar:07} 
take as input white space delimited tokens~\cite{Kulick:10},
consider them as words,
and enumerate all possible solutions. 
This approach has several problems. 
First, a white space delimited token may have 
more than one word, referred to as ``run-on'' 
words~\cite{Buckwalter:04}.
Second, the exhaustive enumeration may not be efficient and may
not be necessary or appropriate
in some applications as noted in~\cite{Maamouri:10}. 
%Other morphological analyzers such as 
%Amira~\cite{Diab:07,Benajiba:07},
%MAGEAD~\cite{Habash:05}, and MADA+TOKAN~\cite{Habash:09} 
%use machine learning and support vector machines (SVM) 
%to disambiguate the morphological analysis at the expense 
%of efficiency. 
%Xerox rules can be compiled into specialized finite state
%machine (FSM) based analyzers as described in~\cite{Beesley:03}.
%However, the efficiency of the resulting analyzer depends on the
%way the Xerox rules are written. 
%This requires deep knowledge and insight from the user
%in compilation techniques, context free grammars, 
%and morphological analysis.
%context free grammar rules that can be automatically compiled
%into efficient non-deterministic finite state Automata (NDFSA).

%\subsection{Sarf}
%\label{sec:intro:sarf}

{\bf Sarf.~~~}
Sarf~\footnote{Sarf (name modified for blind review) }
is 
an {\em application-specific efficient
morphological analyzer} based on
non-deterministic finite state Automata 
driven by an application-specific controller customized 
by the user. 
%Each machine in Sarf takes one letter at a time as input
%and represents a valid morphological analysis.
The application-specific controller prunes false positives
early in the run by making a decision at every input letter
to control the machines.
Sarf builds on and extends the lexicon of 
Buckwalter~\cite{Buckwalter:02} 
with proper and location names extracted from different online 
sources~\footnote{\href{http://alasmaa.net/}{http://alasmaa.net/ }, 
\href{http://ar.wikipedia.org/}{http://ar.wikipedia.org/}}
as well as biblical sources~\footnote{Genesis 4:17-23; 5:1-32; 9:28-10:32; 11:10-32; 25:1-4, 12-18; 36:1-37:2; Exodus 6:14-25; Ruth 4:18-22; 1 Samuel 14:49-51; 1 Chronicles 1:1-9:44; 14:3-7; 24:1; 25:1-27:22; Nehemiah 12:8-26; Matthew 1:1-16; Luke 3:23-38}.

%Each alive machine in Sarf takes as input one letter at a time 
%from a text stream. 
Sarf does not assume that the word at hand is a token and
performs tokenization on the fly based on morphological correctness
and thus can handle 
the ``run-on'' words problem. 
Sarf also uses diacritics if they exist to disambiguate 
morphological solutions.

In this paper we make the following contributions. 
\begin{itemize}
\item We automate the segmentation of hadith books into separate narrations and extract chains of narrators 
%into a partial order relation graph 
using a knowledge based approach with morphological analysis.
\item We optimize our analysis through utilizing punctuation and learning previously unknown names.
\item We build from different chains of narrators a partial order relation graph based on merging common narrators.
\item We introduce a specialized distance metric for Arabic names used in Arabic literature. 
\item We highly outperform the currently available narrator extraction tool. 
\item We provide a working and efficient implementation of our work. 
\end{itemize}

The rest of this paper is organized as follows. 
We describe our approach to narrator chain extraction
in Section~\ref{sec:islamic}.
We describe how we build the partial order relation
graph in Section~\ref{sec:graph}.
We discuss related work in Section~\ref{sec:related}.
We present our results in Section~\ref{sec:results} and 
we conclude in Section~\ref{sec:future}.

%The hadith 
%application specific controller that controls Sarf considers 
%proper names and ignores text where
%proper names occur rarely.
%It also considers words that relate names to each other 
%such as \RL{bin} or ``the son of''
%and words that mean narrate such as \RL{qaal} or ``said''.

%With Sarf, we successfully automated the analysis of 
%three books of Islamic narrations~\cite{IbnHanbal,AlTousi,AlKulayni}
%and extracted from each one of them a vector of complex structures 
%that are three level deep in hierarchy.
%We also merged these vectors into a graph of narrators where 
%we represented identical narrators from separate traditions 
%in one single node.

>>>>

\usepackage{arabtex}
\usepackage{amssymb}
\usepackage{caption}
\usepackage{epsfig}
\usepackage{subfigure}
\usepackage{color}
\usepackage{rotate}
\usepackage{rotating}
%\usepackage{amsthm}
%\usepackage{booktabs}

\usepackage{relsize}
%\usepackage{fancyvrb}
\usepackage[colorlinks=false,draft]{hyperref}
%\usepackage[colorlinks=false]{hyperref}

%\allowarab{\fbox}
\usepackage{utf8}
\setarab
\fullvocalize
\transtrue
\arabtrue

%\newcommand{\CharCodeIn}[1]{`\CodeIn{#1}'}
\newcommand{\true}{\ensuremath{\mathit{true}}}
\newcommand{\false}{\ensuremath{\mathit{false}}}

\newcommand{\cci}[1]{{\small\texttt{#1}}}
\newcommand{\chci}[1]{\cci{#1}}

\newcommand{\frl}[1]{\fbox{\RL{#1}}} 
\newcommand{\noArRL}[1]{\arabfalse\RL{#1}\arabtrue} 
\newcommand{\noTrRL}[1]{\transfalse\RL{#1}\transtrue} 
\newcommand{\noTrnoVocRL}[1]{\transfalse\novocalize\noTrRL{#1}\vocalize\transtrue}  
\newcommand{\noVocRL}[1]{\transtrue\novocalize\RL{#1}\vocalize}  
\newcommand{\utfRL}[1]{\setcode{utf8} \RL{#1} \setcode{standard}}
\newcommand{\noTrnoVocutfRL}[1]{\setcode{utf8} \transfalse\novocalize\RL{#1}\vocalize \transtrue \setcode{standard}}
\newcommand{\noTrutfRL}[1]{\setcode{utf8} \transfalse\RL{#1}\transtrue \setcode{standard}}
\newcommand{\noArutfRL}[1]{\setcode{utf8} \arabfalse \RL{#1} \arabtrue \setcode{standard}}

%\title{Instructions for NAACL HLT 2010 Proceedings\Thanks{This...}}
\title{Detection of Arabic Entity Graphs using Morphology, Finite State Machines, and Graph Transformations}

%\author{ Jad Makhlouta \\\And
%Hamza Harkous \\
%  American University of Beirut \\
%  {\tt \{jem04, hhh20, fz11\}@aub.edu.lb } \\\And 
%  Fadi Zaraket 
%}
\date{}

\begin{document}
\maketitle

\begin{abstract}
Research on automatic recognition of named entities
%events and relations amongst them 
from Arabic text uses techniques that work well for the 
Latin based languages such as local grammars, 
statistical learning models, pattern matching, 
and rule-based techniques. 
These techniques boost their results by using application 
specific 
corpora, parallel language corpora,
and morphological stemming analysis. 
We propose a method for extracting entities, events, 
and relations amongst them
from Arabic text using 
a hierarchy of finite state machines driven by morphological 
features such as part of speech and gloss tags,
and graph transformation algorithms. 
We evaluated our method on three natural language processing 
applications.
We automated the extraction of narrators and 
narrator relations from 
several corpora of Islamic narration books (hadith). 
We automated the extraction of genealogical family trees 
from Biblical texts. 
Finally, we automated locating and extracting
individual biographies from historical biography books.
In all applications, our method reports high precision and recall
and learns lemmas about phrases that improve results. 
\end{abstract}

\section{Introduction}
Named entity recognition (NER) is harder for Arabic text than 
it is for Latin. 
Arabic does not have upper case letters which distinguish the 
start and end of named entities (NE).
Arabic is morphologically rich, 
more ambiguous than Latin languages,
and its short vowels are frequently omitted~\cite{Debili1998}.

Industrial tools that perform NE extraction for Arabic exist~\cite{Sak09,IdentiFinder,basis:06,ANEE:07},
but little is said about the techniques behind them and no formal evaluation of the tools exist. 
Researchers proposed and evaluated 
local grammars with morphological stemming~\cite{ZAGHOUANI10,Traboulsi:09},
rule-based systems powered with local grammars~\cite{ShaalanR09} 
and stop words ~\cite{Abuleil04},
statistical analysis boosted with Arabic-English parallel corpora~\cite{BenajibaZDR10}, 
statistical analysis boosted with task specific gazetteers~\cite{ANERSys:07,BenajibaDR08},
and pattern matching powered with morphological stemming~\cite{jumaily2011,TAGARAB98}
to perform Arabic NER.
We discuss these techniques and compare to them in 
the ``Related work'' Section.

In this paper, we consider the problem of NER
in the context of more complex natural language
processing (NLP) tasks that look for NEs and 
relations amongst the detected NEs and build entity graphs.
An {\em entity graph} is a graph where NEs are nodes and relational entities (RE) are edges.
For example, consider the tasks of extracting a time table
from a document describing the events of a conference, 
extracting genealogical family trees based on person names 
from biblical text,  and
extracting routes and maps from travel itineraries.
Also consider applications with cross-document NER
such as extracting narrator authentication graphs based on narration 
documents that include narrator sequences, and biography documents 
that include credibility information about the narrators. 
We hypothesize that NEs of the same kind happen 
at proximity of each other, and we base our method on that 
working hypothesis.
We target NLP applications and corpora where our working hypothesis
obviously holds.
These applications help build databases of NEs
that we can later use in other related contexts.

We propose a method for extracting NEs and 
relations amongst entities using 
morphological features such as part of speech (POS) and gloss tags, 
a hierarchy of finite state machines, 
and graph transformations such as node merge, node split, and replace edge transformations. 
Morphological analysis helps reduce ambiguities and resolves lexical items with their morphological variations.
For each application, we augment the lexicon of an inhouse morphological 
analyzer with lexical elements and tags that describe categories relevant to the application. 
We also build finite state transducers (FST) that take 
morphological tags as input and detect the NEs.
Manually built FSTs can express rules that are not intuitive (and sometimes impossible) 
to express with local grammars using regular expressions.
Consider, for example, counting the number of tags up to a threshold, or capturing left and right recursive structures 
which are common in Arabic since an Arabic phrase-sentence can start with a name or a verb.
We finally infer relations between the NEs based on their context.

Our method takes text and computes morphological features that match, precede, follow, or connect NEs
and uses the features to learn unknown NEs.
Our method also computes morphological features that correspond to relations between NEs, and learns semantics
for them. 
For example, consider a graph with two predefined edge labels $l_s=$``spouse'' and $l_p=$``parent'', 
with names of persons $n_1, n_2,$ and $n_3$ as nodes,
and with $(n_1,n_2,l_s)$, $(n_2,n_3,l_p)$, and $(n_1,n_3,l_p)$ as edges inferred from a document.
Our method can learn that the expression \utfRL{ولدت له} (gave him a child) connecting 
$n_1$ to $n_2$ and $n_3$ means that $n_2$ is the spouse of $n_1$
and the parent of $n_3$.
The learned semantics for the expression can be expressed as a graph transformation that leads to a graph with only 
$l_s$ and $l_p$ as labels.

We evaluated our method and used it to solve three 
NLP applications including extracting genealogical family
trees from Biblical text, extracting a graph of 
narrators from hadith books, and extracting biographies from biography books.
Our results show high precision and recall metrics and report
learning interesting named entities and relational lemmas.
In this paper we make the following contributions.
\begin{itemize}
\item We present a method for named entity and entity graph
extraction from Arabic text using morphology, state 
machines, and graph transformation algorithms.
\item We present an Arabic lexicon augmented 
with (1) person names collected from several resources as well as
learned from our corpora, and (2) tags suitable to extract complex names,
kinship, and narration sequence relations. 
\item We learn semantics for morphological features associated 
with detected phrases that are equivalent to predefined features
modulo one or two graph transformations. 
The learned semantics improved recall and precision by 
5 and 6\%, respectively in the genealogy application.
\item We perform Arabic cross-document named entity and 
relational entity extraction using graph algorithms. 
The use of a graph generated from the hadith application 
improved precision in the biography application from 40\% to 93\%.
%\item We present a novel equality metric to compare Arabic 
%names based on name structure and morphological features. 
\end{itemize}

\section{The method}

Our method takes a set of Arabic text documents and builds 
a graph where named entities are nodes and edges are relations 
between the nodes. 
Our target graph has a predefined set of relations expressed 
as edge labels such as ``parent'' and ``spouse'' where the nodes
are person names.
We use an inhouse morphological analyzer augmented with 
a list of lexicon elements relevant to the application.
We also augment the analyzer with
a set of tags that classify the 
elements into several categories 
relevant to the targeted named entities and edge labels.

The method passes the input text string to the morphological 
analyzer. The analyzer processes the input text and whenever
it identifies a morpheme, it calls the application morpheme
classifier with the morpheme, the tags of the morpheme, 
and the current solution context.
Note that several morphological solutions may exist for the same
input string.
The morpheme classifier either stores the morpheme in a list of
unresolved morphemes, 
or resolves the unresolved morphemes and produces
a category as input to the application FST.
The application FST is manually built to
detect {\bf (1)} named entities,  
{\bf (2)} sequences of named entities,  
{\bf (3)} predefined edges connecting named entities, 
{\bf (4)} previously unknown named entities occurring between predefined edges, 
and {\bf (5)} previously unknown edge labels occurring between named entities.
%Note that there might be more than one level of finite state
%transducers to handle an NLP application.
The detected named entities and edges form an entity graph.

Finally, the method uses a morphological and structural distance
function to compute equality between the named entities and
to consequently merge the nodes of the graph that represent 
equal named entities.
%When nodes are merged, the edge labels are augmented to be the 
%concatenation of old edges.
The method then learns semantics for the discovered edge labels 
and suggests edge replacement algorithms to transform the 
discovered edge labels into the predefined edges. 
The method also interactively learns the semantics of other 
detected edge labels that it could not infer 
in terms of the predefined set of relations.

%-importance of cluster (Min number of entities in context)
%-importance of tolerance 
%-entities using lexicons
%-clusters
%-

\section{The hadith application}
A \RL{.hady_t} (tradition) 
is a narration related to the prophet Mohammad
through a \RL{sanad} or a sequence of narrators. 
The collections of traditions are the second source of
jurisprudence after the \RL{qor'An} for all Islamic 
schools of thought. 
Figure~\ref{f:exhadith} shows an example \noArRL{.hady_t} in 
Arabic with its transliteration and translation. 
We show proper names in boxes connected
to form complex names of narrators. 
For example, 
\novocalize
\noTrRL{qtybT} is the first name 
of narrator $n_1$, and 
\noTrRL{s`yd} is the name of his father as 
the word \noTrRL{bn} (son of) indicates. 
The sequence of names from $n_1$ to $n_5$ 
constitutes the \noArRL{sanad} 
of the \noArRL{.hady_t}. 
The second part of the \noArRL{.hady_t} is the 
\RL{matn} (content) and constitutes its actual content.

Due to religious and political reasons, writing the traditions was forbidden 
until the days of the Umayad Calif,
\novocalize
\RL{`mr bn `bd al`zyz}(717-720 AC), seventy or so years after the prophet. 
Many inconsistencies were introduced to the literature which necessitates 
a thorough authentication study before the use of a tradition in jurisprudence.
\vocalize
%An Islamic hadith scholar studies the authenticity of 
%the \noArRL{sanad} of a 
%set of related traditions before using these traditions 
%jurisprudence. 
%While different Islamic schools of thought differ on 
%how to interpret the content, they almost all agree
All Islamic schools of thought agree
that scholars can not use a tradition whose sanad lacks authenticity for jurisprudence.

The authenticity of a \noArRL{.hady_t} depends on 
the credibility of the narrators as reported in 
separate biography books. 
The study of \noArRL{.hady_t} authentication is 
currently manual and error prone due to the huge number
of existing traditions and tradition books. 
%Hadithopedia~\cite{Hadithopaedia:08}
%reports that a non-comprehensive collection of tradition 
%and biography books amounts to more than 300 thousand lines of 
%text. 
Al-Azami\cite{Al-Azami-91} cites more than eleven books
of digitized tradition books each of several volumes, and a dozen
other biography and secondary authentication books such
as a geographical dictionary of places in hadith.

%In this paper, we present the hadith {\em narrator graph
%extractor using morphological analysis} (NaGEMA), 
%a novel technique that automates
%extracting the \noArRL{sanad} from the books of traditions 
%into a diagram of sequences of narrators. NaGEMA segments
%a book into narrations, each narration into its \noArRL{matn} 
%and its \noArRL{sanad}, 
%and each sanad into its separate narrators. 
%This work will lead to 
%automate an exhaustive \noArRL{.hady_t} authentication 
%effort in the future as we plan to 
%automatically analyze the biography books.

%Automated analysis of Arabic data sets, including texts, 
%publications, records and digital media is essential
%with the huge digital Arabic content available nowadays. 
%\subsection{ Arabic morphological analysis}
%Arabic morphological analysis is key to our analysis. 
%Current morphological analyzers~\cite{Sughaiyer:04}
%use concatenative analysis when
%considering the internal structure 
%of an Arabic word and
%composing it into several {\em morphemes}. 
%A morpheme can be a {\em stem}, or an {\em affix}.
%An affix can be a {\em prefix, suffix, } or an {\em infix}.
%%The analysis of one word may lead to several possible
%%morphological solutions.
%\vocalize
%The word \RL{'a.hmadH}
%may have two valid morphological analyses. 
%The letter \RL{'a} may be a prefix and the word means 
%``I praise him'', or 
%%The letter \RL{'a} may also be 
%part of the stem \RL{'a.hmad} (a proper noun)
%and the word means ``his Ahmad''.
%
%The morphological analysis of the word \RL{.hd_tnA} 
%``narrated to us'' that starts the hadith in Figure~\ref{f:exhadith}
%returns the stem \RL{.hd_t} ``narrated''.
%The stem  \RL{.hd_t} is also the stem of other 
%words such as \RL{.hd_tny} ``narrated to me'' and \RL{.hd_thm} ``narrated to them''.
%The stem also shares similar part of speech (POS)
%and English meaning gloss annotations with words
%such as \RL{qaal} ``said'' and \RL{'a_hbar} ``told''. 
%The use of morphological analysis allows us to 
%subsume listing these stems and all of their morphological
%derivations with one category that matches the gloss and 
%POS tags. 
%Similarly the words \RL{ibn} ``the son of'' and \RL{bn} 
%return the same morphological analysis.
%
%
%\subsection{Narrator sequence extraction }

%\begin{figure*}
%\center{
%\resizebox{.9\columnwidth}{!}
%{\includegraphics{figs/detailed_submachine.pdf}
%}
%\caption{The sequence of narrators extractor diagram}
%\label{f:sequence}
%}
%\end{figure*}

@@

\transfalse
\begin{figure}[tb]
\center{
\resizebox{.95\columnwidth}{!}
{ \input{figs/exhadith2cols.pdftex_t}}
\caption{Hadith narrator sequence example.}
\label{f:exhadith}
}
\end{figure}
\transtrue
\vocalize

Several researchers~\cite{iTree,Hadithopaedia:08} attempted to automate 
the analysis of the hadith literature. 
We fully automate the analysis of three books of hadith selected 
arbitrarily~\cite{IbnHanbal,AlKulayni,AlTousi}~\footnote{We obtained
  the digitized books from 
  \href{http://www.yasoob.com/}{http://www.yasoob.com/} and 
  \href{http://www.al-eman.com/}{http://www.al-eman.com/}. }.
We accept a book as input
and segment it into a vector of hadiths,
and segment each hadith into its sanad and matn parts. 
We detect the sequence of narrators in the sanad and 
the relation that links each narrator to his ancestor and 
predecessor in the sequence. 
We also detect the full name of each narrator that is
composed of several proper names with connectors
in between.

%NRC=19 descriptions= 91 stems
%family nmc connectors (all)=15 descriptions=33stems
%       ibn=1 descriptions=3 stems
%       2ab =2 descriptions=11 stems
%       2om=4 descriptions =5stems

To do that, we augmented our lexicon with person names obtained from online and Biblical sources,
and augmented the tag set of our lexicon with the following categories.
\begin{itemize}
\item $\mathit{NAME}$: a proper name of a person such as \RL{'a.hmad}. (2,500 male and 3,327 female lexicon items). 
\item $\mathit{IBN}$: a special name connector meaning ``the child of'' such as \RL{ibn}. (3 lexicon items expressing child,
11 expressing father, and 5 expressing mother relations).
\item $\mathit{NMC}$: name connectors that may be any word appearing amongst the other name connector categories. (33 lexicon items).
\item $\mathit{NISBA}$: a possessive adjective that qualifies a person such as \RL{alma.sriy} ``the Egyptian''. (4,698 possessive city or town, 82 country, and 4 continent items). 
\item $\mathit{NRC}$: a narrator connector such as
\RL{`an} ``on behalf of'', \RL{.hada_t} ``narrated'', \RL{qaal} ``said'', \RL{'a_hbar} ``told'' or one of their morphological equivalents. (91 lexicon items).
\end{itemize}
The categories can also be inferred from the concatenation of the tagged morphemes.

\subsection{The hadith finite state transducer}
\label{sec:FST}
% FSM for the FST

We implemented our hadith sequence extraction 
in an FST that reports a valid sequence
of narrators when a sequence of names
connected by narrator connectors appears. 
%This includes the task of finding multi-word names
%such as \RL{`bd alr.hmn} often appearing as ``run-on'' words.
The FST marks the beginning and the end of the hadith with 
the beginning of the current name sequence
and the beginning of the next name sequence, respectively.
%It detects the sequence of narrators as the sequence itself.

%The FST also targets words that mean ``narrate'' when
%they appear between proper names. 
We illustrate the hadith FST
in Figure~\ref{f:hadith}(a). 
The transitions are  labeled with expressions
that operate on the category tags 
generated from the morphological analyzer.
The FST has four abstract states.
State $\mathit{TEXT\_S}$ is the initial state and denotes 
that the FST is outside the context of a sanad.
State $\mathit{NAME\_S}$ denotes that the FST is in
the middle of a sequence and has met a name. 
State $\mathit{NMC\_S}$ denotes that the FST is in
the middle of a narrator name.
State $\mathit{NRC\_S}$ denotes that the FST is 
in the context of a narrator name and the FST 
has met a narrator connector and is ready to meet 
a new narrator.

The threshold $\theta_{\mathit{nmc}}$ 
corresponds to the number of tolerated name connectors 
that may occur between two names. %3
The symbol $\mathit{LIST}_{\mathit{nmc}}$ corresponds to the list 
of name connectors collected since the FST
started looping in the state $\mathit{NMC\_S}$.
The symbol $\lambda_{\mathit{nmc}}$ is a parameter 
that corresponds to a relaxed tolerance measure that
the FST resorts to in case the words separating
two names were longer than $\theta_{\mathit{nmc}}$ but 
contained a name connector word such as $\mathit{IBN}$ 
or $\mathit{NISBA}$.

The FST moves to state $\mathit{NAME\_S}$ on
$\mathit{NAME}$.
It moves to state $\mathit{NRC\_S}$ 
when Sarf reports an $\mathit{NRC}$.
State $\mathit{NMC\_S}$
indicates that the FST expects a name to appear within 
a tolerance threshold expressed by 
$\theta_{\mathit{nmc}}$ and $\lambda_{\mathit{bmc}}$.
It returns to $\mathit{NAME\_S}$ when a name is met, 
loops when still within the threshold, and 
resets to $\mathit{TEXT\_S}$ when the threshold is exceeded 
signaling that the last collection of names met do not qualify
as a narrator or a sequence of narrators.

Similarly, $\mathit{NRC\_S}$ 
tolerates $\theta_{\mathit{nrc}}$ words 
before it gives up on its expectations. 
Note that we reach the $\mathit{NAME\_S}$
state only when a
valid $\mathit{NAME}$ is detected and we leave when no 
more $\mathit{NAME}$'s are detected.
The $\mathit{NRC\_S}$ state can only be reached if an 
$\mathit{NRC}$ is detected.
The state $\mathit{NMC\_S}$ can only be reached from a 
$\mathit{NAME\_S}$ state.

Whenever the FST transitions to $\mathit{NRC\_S}$ from either
$\mathit{NAME\_S}$ or $\mathit{NMC\_S}$, the FST reports 
a narrator detected and adds it to $\mathit{LIST}_{\mathit{narr}}$.
Whenever the FST transitions back to $\mathit{TEXT\_S}$,
the FST reports a sanad detected if 
$|\mathit{LIST}_{\mathit{narr}}| \ge N_{min}$ where 
$\mathit{LIST}_{\mathit{narr}}$ is the list 
of detected narrators and $N_{min}$ is a threshold denoting 
the minimum number of narrators required in a sanad.

We used the values of 3, 5 and 5 for $\theta_{\mathit{nmc}}$, 
$\lambda_{\mathit{nmc}}$, and 
$\theta_{\mathit{nrc}}$, respectively to obtain the results in 
Table~\ref{t:hadithresults}.

\begin{figure}[tb!]
\center{
\begin{tabular} {cc}
\resizebox{.45\columnwidth}{!}
{ \input{figs/hadith.pdftex_t}} 
&
\resizebox{.55\columnwidth}{!}
{\includegraphics{figs/DAG_narrator.pdf}} \\
(a) FST & (b) Extracted narrator graph 
\end{tabular}
\caption{Hadith application.}
\label{f:hadith}
}
\end{figure}

%\begin{figure}[tb!]
%\center{
%\resizebox{.6\columnwidth}{!}
%{ \input{figs/hadith.pdftex_t}}
%\caption{The FST of the hadith application.}
%\label{f:hadith}
%}
%\end{figure}

%The definition of input labels such as NAME and IBN depends on the 
%morphological analyzer. 
%However, our case-based FST approach was able to perform well
%under both Sarf and a refined version of Sarf.

%\subsection{Optimizations}
%
%In addition to the FST, 
%NaGEMA uses punctuation marks to refine its analysis and 
%increase the accuracy of name detection, narration detection and narrator boundaries. 
%It also learns narrator names from simple patterns that 
%happen within narrator connectors and that are not 
%reported by Sarf as names. 
%NaGEMA learns names of narrators that consist of words missing from 
%the Sarf lexicon.
%NaGEMA learns the narrator names based on their context as they occurred 
%between narrator connectors or contained name connectors. 
%
%(explain better and merged information present about them in "Results" section)
%
\subsection{Extracting the narrator graph using node merge and split transformations}
\label{sec:graph}

%\begin{figure}[tb]
%\center{
%\resizebox{.8\columnwidth}{!}
%%{\includegraphics{figs/narrator_sequence_output.pdf}}
%{\includegraphics{figs/DAG_narrator.pdf}}
%\caption{Narrator directed acyclic graph 
%    extracted using NaGEMA }
%\label{f:narrators} 
%}
%\end{figure}
